|

### 2024-05-21 Drafts/Paper_1/paper_1.Rmd
mid-edit update

@@

  save(years.tmp,pnm.tmp,waa.tmp,ages.tmp,rem.tmp,mx.tmp,NE.tmp,vpa.tmp,mr.tmp,
       am.tmp,Stocks,ASR.stocks,ASR.long,abund.tmp,fs.tmp,full.rem.tmp,
       file =  paste0(loc,"/Results/model_inputs_no_age_correction.Rdata"))

@@

  # Doubling time, no fishing
  dt.res <- NULL
  # These are both set in the index file now, go there!!
  n.dt.sims <- n.dt.sims
  n.sim.years <- n.sim.years
  for(i in Stocks)
  {
    str.time <- Sys.time()
    try <- for.tune.all[[i]]
    meta.tmp <- try$meta
    age.4.sim <- meta.tmp$age.min:meta.tmp$age.max
    double.time <- NA
    pop <- 1000
    tmp.res <- NULL
  for(s in 1:n.dt.sims)
  {
    tmp <- rep(0,n.sim.yearss)
    print(paste0("Sim # ", s))

    for(y in 2:n.sim.yearss)
    {

      # Get our sample for this run
      f.pick <- sample(nrow(try$fecund.opt),size=1)
      nm.pick <- sample(nrow(try$nm.opt),size=1)
      fec.samp <- try$fecund.opt[f.pick,]
      # This is instantaneous rate, which is what the Lotka function is expected (it is set up expecting instantaneous)
      nm.samp <- try$nm.opt[nm.pick,] 
      # now put these into Euler-lotka to get an r estimate
      res <- simple.lotka.r(mort = nm.samp,fecund=fec.samp,ages=age.4.sim)
      r.val <- res$res$r
      # If r is > 5 (lambda = exp(5)) it generally means the simple.lotka didn't converge (bad combo of fec and nm)
      # So we'll need to resample them until it is < 5
      while(r.val >= 5)
      {
        print("Resampling due to bat shit r value")
        f.pick <- sample(nrow(try$fecund.opt),size=1)
        nm.pick <- sample(nrow(try$nm.opt),size=1)
        fec.samp <- try$fecund.opt[f.pick,]
        # This is instantaneous rate, which is what the Lotka function is expected (it is set up expecting instantaneous)
        nm.samp <- try$nm.opt[nm.pick,] 
        # now put these into Euler-lotka to get an r estimate
        res <- simple.lotka.r(mort = nm.samp,fecund=fec.samp,ages=age.4.sim)
        r.val <- res$res$r
      } # end while loop
      pop[y] <- exp(r.val)*pop[y-1]
      if(pop[y] >= 2*pop[1] & tmp[y] == 0) tmp[y:n.sim.yearss] <- 1
    } # end the years loop
    tmp.res[[s]] <- data.frame(year =1:n.sim.yearss,doubled = tmp,sim=s)

@@

# This extracts the GAM fit so we can custom plot it on our own figure
tmp <- plot(mod.dd.stock.gam.nf, pages = 0)
no.stock.gam.nf.dat <- data.frame(fit = tmp[[1]]$fit + coefficients(mod.dd.stock.gam.nf)[1],prop.abund = tmp[[1]]$x,se = tmp[[1]]$se,Scenario = "F = 0")
no.stock.gam.nf.dat$LCI <- no.stock.gam.nf.dat$fit - 1.96*no.stock.gam.nf.dat$se
no.stock.gam.nf.dat$UCI <- no.stock.gam.nf.dat$fit + 1.96*no.stock.gam.nf.dat$se

@@

tmp <- plot(mod.dd.stock.gam.fish, pages = 0)
no.stock.gam.fish.dat <- data.frame(fit = tmp[[1]]$fit + coefficients(mod.dd.stock.gam.fish)[1],prop.abund = tmp[[1]]$x,se = tmp[[1]]$se,Scenario = "F > 0")
no.stock.gam.fish.dat$LCI <- no.stock.gam.fish.dat$fit - 1.96*no.stock.gam.fish.dat$se
no.stock.gam.fish.dat$UCI <- no.stock.gam.fish.dat$fit + 1.96*no.stock.gam.fish.dat$se

@@

# the lambda calcs
med.lambda.F0 <- round(median(res.lambda.fin$lam.no.fish),digits=3)
med.r.F0 <- round(median(log(res.lambda.fin$lam.no.fish)),digits=3)*100
med.lambda.F <- round(median(res.lambda.fin$lam.fish),digits=3)
med.r.F <- round(median(log(res.lambda.fin$lam.fish)),digits=3)*100
mn.lambda.F <- round(mean(res.lambda.fin$lam.fish),digits=2)
mn.r.F <- round(mean(res.lambda.fin$lam.fish),digits=2)
# Mean - median...
res.lambda.fin$med.mn.diff.no.fish <- res.lambda.fin$lam.no.fish.med - res.lambda.fin$lam.no.fish.mn
res.lambda.fin$med.mn.diff.fish <- res.lambda.fin$lam.fish.med - res.lambda.fin$lam.fish.mn
# How many stocks had the mean higher than the median
n.mn.higher.med.no.fish <- length(unique(res.lambda.fin$Stock.short[res.lambda.fin$med.mn.diff.no.fish < 0]))
n.mn.higher.med.fish <- length(unique(res.lambda.fin$Stock.short[res.lambda.fin$med.mn.diff.fish < 0]))
low.mn.lambda.stocks <- unique(res.lambda.fin$Stock.short[res.lambda.fin$med.mn.diff.fish > 0])
n.nw.ice <- length(c(grep("NAFO",low.mn.lambda.stocks),grep("Faroe",low.mn.lambda.stocks)))
n.cod.low.mn <- length(grep("morhua",low.mn.lambda.stocks))
n.cod <- length(grep("morhua",Stocks))

@@

>>>>

---
title: |
   "Life history variability influence on the recovery potential of globally exploited teleosts "
french_title: "Titre ici (*Nom latin de l'espèce*)"
author: |
  David. M. Keith^1^ 
  Freya M. Keyser^1^
  Julie M. Charbonneau^2^
  Heather 'The Boss' Bowlby^1^
author_list: "Keith, D.M., Keyser, F.M., Charbonneau, J.M., Bowlby, H.T.B."
address: |
  ^1^Bedford Institute of Oceanography\
     Fisheries and Oceans Canada, 1 Challenger Dr.\
     Dartmouth, Nova Scotia, B2Y 4A2 , Canada\
  \smallskip
  ^2^Simon Fraser University\
     Department of Biology\
french_address: |

month: ""
french_month: ""
year: 2024
report_number: nnn
region: "Maritimes Region"
french_region: "Maritimes Region"
isbn: ""
cat_no: ""
citation_other_language: ""

abstract: |
  Here is the abstract text. Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.

  Start new paragraphs after a blank line and with 2 spaces indent. Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.

french_abstract: |

header: "Draft working paper --- Do not cite or circulate" # or "" to omit
output:
 csasdown::resdoc_word:
   french: false
   copy_sty: true
   line_nums: true
   line_nums_mod: 1
   lot_lof: false
   draft_watermark: false
   include_section_nums: true
   highlight: tango
# ------------
# End of options to set
knit: (function(input, ...) {
       csasdown::render('_bookdown.yml')
      })
link-citations: true
bibliography: Y:/Zotero/MAR_SABHU.bib
# Any extra LaTeX code for the header:
# header-includes:
# - \usepackage{tikz}
---

```{r setup, echo=FALSE, cache=FALSE, message=FALSE, results='hide', warning=FALSE}
library(knitr)
if (is_latex_output()) {
  knitr_figs_dir <- "knitr-figs-pdf/"
  knitr_cache_dir <- "knitr-cache-pdf/"
  fig_out_type <- "png"
} else {
  knitr_figs_dir <- "knitr-figs-docx/"
  knitr_cache_dir <- "knitr-cache-docx/"
  fig_out_type <- "png"
}
fig_asp <- 0.618
fig_width <- 9
fig_out_width <- "6in"
fig_dpi <- 180
fig_align <- "center"
fig_pos <- "htb"
opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  comment = "#>",
  fig.path = knitr_figs_dir,
  cache.path = knitr_cache_dir,
  fig.asp = fig_asp,
  fig.width = fig_width,
  out.width = fig_out_width,
  echo = FALSE,
  #  autodep = TRUE,
  #  cache = TRUE,
  cache.comments = FALSE,
  dev = fig_out_type,
  dpi = fig_dpi,
  fig.align = fig_align,
  fig.pos = fig_pos
)
```

```{r load-libraries-n-data, echo = FALSE, cache = FALSE, message = FALSE, results = 'hide', warning = FALSE}
# add other packages here:
library(dplyr)
library(readr)
library(tibble)
library(csasdown)
# OK, so using the ICES assessments here's what we get for North Sea cod.
library(readxl)
library(tidyverse)
library(rio)
library(ggthemes)
library(cowplot)
library(gtable)
library(viridis)
library(arm)
library(mgcv)
library(lme4)
library(scales)
library(ggh4x)
library(gratia)
theme_set(theme_few(base_size = 14))

# The location of your data
loc <- 'D:/GitHub/ICM'
# The parameters for the doubling time simulation
n.dt.sims <- 1000 # Increase to 1000 for final run, which takes like a full day or something 
n.sim.years <- 100
# If we already have the results don't re-run the models or load data we don't need.
load.results <- F
# What tunning model are you running or did you run!
what.2.tune <- 'fec_nm'

#source(paste0(loc,"/Scripts/functions/tuning_sim_fast.R"))

# Download the functions we'll need from github
funs <- c("https://raw.githubusercontent.com/dave-keith/ICM/main/Scripts/functions/simple_Lotka_r.r",
          "https://raw.githubusercontent.com/dave-keith/ICM/main/Scripts/functions/backwards_project.r",
          "https://raw.githubusercontent.com/dave-keith/ICM/main/Scripts/functions/forward_project.r",
          "https://raw.githubusercontent.com/dave-keith/ICM/main/Scripts/functions/tuning_sim_fast.r"
)
  # Now run through a quick loop to load each one, just be sure that your working directory is read/write!
for(fun in funs) 
{
  download.file(fun,destfile = basename(fun))
  source(paste0(getwd(),"/",basename(fun)))
  file.remove(paste0(getwd(),"/",basename(fun)))
}

if(load.results==F)
{
  # Choose 5 ICES stocks that we have the necessary data for
  #ASR <- read_xlsx("../Data/ASR_2018.xlsx" sheet = "ICES")
  ASR1 <- read.csv(paste0(loc,"/Data/ICM_data_NE_atlantic_stocks.csv"))
  ASR <- read.csv(paste0(loc,"/Data/ICM_data_NW_atlantic_stocks.csv"))

  ASR <- rbind(ASR1,ASR)
  datatypes <- unique(gsub(x = names(ASR), pattern = "[^a-zA-Z]", replacement=""))

  # Replace 0's in Num.tot with NA so the rest of this works as Freya designed it to...
  ASR$Num.tot[ASR$Num.tot == 0] <- NA
  # we want:
  # Year, Num, WA, Catch, AM, NM, StockID, Management, Area, Order, Family, Genus, Species
  ASRdat <- ASR[,c(grep(x=names(ASR), "Num"),
                   grep(x=names(ASR), "WA"),
                   grep(x=names(ASR), "Catch"),
                   grep(x=names(ASR), "AM"),
                   grep(x=names(ASR), "NM"))]
  # Making all the data numeric that should be numeric
  ASRdat <- apply(X = ASRdat, 2, as.numeric)
  # Getting the species info back
  ASRsp <- ASR[, which(!1:length(names(ASR)) %in% grep(x=names(ASR), ".", fixed=T))]
  # And binding it all back together
  ASR.trim <- cbind(ASRsp, ASRdat)
  # need a unique ID for stock
  #table(ASR.trim$Management, ASR.trim$Species)
  ASR.trim$Stock <- paste0(ASR.trim$Management, "_", ASR.trim$Area, "_", ASR.trim$Genus, "_", ASR.trim$Species)

  ASR.long <- ASR.trim %>%
    pivot_longer(!c("Management", "Area", "Order", "Family", "Genus", "Species", "Stock", "Year","Meeting_or_reference","Model","Case","Notes")) %>%
    separate(col=name, into=c("type", "age"), sep = "\\.")

  # Something is weird with the mentella stock, I'm going to drop it...
  ASR.long <- ASR.long %>% dplyr::filter(Stock != "ICES-AFWG_DEEP1-2_Sebastes_mentella")
}# end if load results
```

```{r bilingual-code, cache=FALSE}
options(
  # Prevent xtable from adding a timestamp comment to the table code it produces
  xtable.comment = FALSE,
  # Do not allow kableExtra to load packages, we add them manually in csasdown
  kableExtra.latex.load_packages = FALSE,
  # Stop chunk output (echo) running into the margins
  width = 80,
  # Do not use scientific notation (stops tables from showing 1.2e3, etc.)
  scipen = 999)

meta <- rmarkdown::metadata
meta_out <- rmarkdown::metadata$output
csl <- "csl/csas.csl"
options(OutDec = ".")
if (is.null(getOption("french"))) {
  stop("`french` was not set up correctly in YAML header in index.Rmd. ",
       "It must be true or false",
       call. = FALSE)
}
if (getOption("french")) {
  csl <- "csl/csas-french.csl"
  options(OutDec = ",")
}

french <- isTRUE(getOption("french")) # for backwards compatibility

# This hook simplifies document translation for the author.
# When building in French, it draws a box around paragraphs contained in chunks
#  which have the option `needs_trans = TRUE`. It also labels
#  the box with a "Needs translation" tag in red and the chunk label in blue.
# You need to change the `needs_trans` option to `FALSE` for a chunk once you
#  have inserted the translated text into it. You will get a utf-8 error if
#  you leave it as `TRUE` and there is French in the chunk.
#  This function assumes you are translating from English to French.
#  If you wrote your document in French and want to translate to English,
#  put a ! before `getOption("french")` below and add the `need_trans`
#  chunk options to your English paragraph chunks instead of the French ones.
#  
#  IMPORTANT NOTES
#  - Use `csasdown::render()` to render the document. This runs a
#    pre-processing step which ensures any inline R chunks
#    `r print("Like this one")` are taken care of correctly and that all
#    backslash variables (eg. \pi, \alpha, \@ref, \cite) are all processed
#    correctly.
#  - French latex places a space before the colon by default so if you need a colon
#    with no space before it, use \\hc .
knit_hooks$set(needs_trans = function(before, options){
  if(getOption("french") && options$needs_trans){
    if (before){
      paste0("\\
   \\begin{lrbox}{\\userinput}
   \\begin{minipage}{\\dimexpr\\linewidth-2\\fboxsep-2\\fboxrule}
   \\textcolor{red}{\\textbf{Needs translation - \\textcolor{blue}{knitr chunk: ", options$label, "}}}
   \\begin{lstlisting}
  ")
    } else {
      "
   \\end{lstlisting}
   \\end{minipage}
   \\end{lrbox}
   \\noindent
   \\fbox{\\usebox{\\userinput}}
  "
    }
  }
})
```

---
title: `r ifelse(fr(), meta$french_title, meta$title)`
month: `r ifelse(fr(), meta$french_month, meta$month)`
region: `r ifelse(fr(), meta$french_region, meta$region)`
csl: `r csl`
---

<!--chapter:end:tmp-index.Rmd-->

@@

  save(years.tmp,pnm.tmp,waa.tmp,ages.tmp,rem.tmp,mx.tmp,NE.tmp,vpa.tmp,#mr.tmp,
       am.tmp,Stocks,ASR.stocks,ASR.long,abund.tmp,fs.tmp,full.rem.tmp,
       file =  paste0(loc,"/Results/model_inputs_no_age_correction.Rdata"))

@@

  # Doubling time, no fishing
  dt.res <- NULL
  # These are both set in the index file now, go there!!
  n.dt.sims <- n.dt.sims
  n.sim.years <- n.sim.years
  for(i in Stocks)
  {
    str.time <- Sys.time()
    try <- for.tune.all[[i]]
    meta.tmp <- try$meta
    age.4.sim <- meta.tmp$age.min:meta.tmp$age.max
    double.time <- NA
    pop <- 1000
    tmp.res <- NULL
  for(s in 1:n.dt.sims)
  {
    tmp <- rep(0,n.sim.years)
    print(paste0("Sim # ", s))

    for(y in 2:n.sim.years)
    {

      # Get our sample for this run
      f.pick <- sample(nrow(try$fecund.opt),size=1)
      nm.pick <- sample(nrow(try$nm.opt),size=1)
      fec.samp <- try$fecund.opt[f.pick,]
      # This is instantaneous rate, which is what the Lotka function is expected (it is set up expecting instantaneous)
      nm.samp <- try$nm.opt[nm.pick,] 
      # now put these into Euler-lotka to get an r estimate
      res <- simple.lotka.r(mort = nm.samp,fecund=fec.samp,ages=age.4.sim)
      r.val <- res$res$r
      # If r is > 5 (lambda = exp(5)) it generally means the simple.lotka didn't converge (bad combo of fec and nm)
      # So we'll need to resample them until it is < 5
      while(r.val >= 5)
      {
        print("Resampling due to bat shit r value")
        f.pick <- sample(nrow(try$fecund.opt),size=1)
        nm.pick <- sample(nrow(try$nm.opt),size=1)
        fec.samp <- try$fecund.opt[f.pick,]
        # This is instantaneous rate, which is what the Lotka function is expected (it is set up expecting instantaneous)
        nm.samp <- try$nm.opt[nm.pick,] 
        # now put these into Euler-lotka to get an r estimate
        res <- simple.lotka.r(mort = nm.samp,fecund=fec.samp,ages=age.4.sim)
        r.val <- res$res$r
      } # end while loop
      pop[y] <- exp(r.val)*pop[y-1]
      if(pop[y] >= 2*pop[1] & tmp[y] == 0) tmp[y:n.sim.years] <- 1
    } # end the years loop
    tmp.res[[s]] <- data.frame(year =1:n.sim.years,doubled = tmp,sim=s)

@@

# This extracts the GAM fit so we can custom plot it on our own figure
tmp <- plot(mod.dd.stock.gam.nf, pages = 0) # Leave a space after this so the code runs (it needs a return as this thing makes a two panel figure)

no.stock.gam.nf.dat <- data.frame(fit = tmp[[1]]$fit + coefficients(mod.dd.stock.gam.nf)[1],prop.abund = tmp[[1]]$x,se = tmp[[1]]$se,Scenario = "F = 0")
no.stock.gam.nf.dat$LCI <- no.stock.gam.nf.dat$fit - 1.96*no.stock.gam.nf.dat$se
no.stock.gam.nf.dat$UCI <- no.stock.gam.nf.dat$fit + 1.96*no.stock.gam.nf.dat$se

@@

tmp <- plot(mod.dd.stock.gam.fish, pages = 0) # Leave a space after this so the code runs (it needs a return as this thing makes a two panel figure)

no.stock.gam.fish.dat <- data.frame(fit = tmp[[1]]$fit + coefficients(mod.dd.stock.gam.fish)[1],prop.abund = tmp[[1]]$x,se = tmp[[1]]$se,Scenario = "F > 0")
no.stock.gam.fish.dat$LCI <- no.stock.gam.fish.dat$fit - 1.96*no.stock.gam.fish.dat$se
no.stock.gam.fish.dat$UCI <- no.stock.gam.fish.dat$fit + 1.96*no.stock.gam.fish.dat$se

@@

# the lambda calcs
n.stock.dfo <- meta.sub |> collapse::fsubset(Manage.short == "DFO-TRAC") |> nrow()
n.stock.ices <- meta.sub |> collapse::fsubset(Manage.short == "ICES") |> nrow()
n.stock.noaa <- meta.sub |> collapse::fsubset(Manage.short %in% c("AFSC","NEFSC")) |> nrow()
n.age.0 <- meta.sub |> collapse::fsubset(age.min == 0) |> nrow()
n.age.1 <- meta.sub |> collapse::fsubset(age.min == 1) |> nrow()
n.age.2.plus <- meta.sub |> collapse::fsubset(age.min >= 2) |> nrow()
min.year <- min(res.lambda.fin$year)
max.year <- max(res.lambda.fin$year)
stock.ts.len <- res.lambda.fin |> collapse::fgroup_by(Stock.short) |> collapse::fsummarise(n.years = length(year))
mn.ts.len <- round(mean(stock.ts.len$n.years),digits=1)
mn.ts.len <- min(stock.ts.len$n.years)
mx.ts.len <- max(stock.ts.len$n.years)
num.orders <- length(unique(meta.sub$Order))
num.gads <- meta.sub |> collapse::fsubset(Order == "Gadiformes") |> nrow()
num.pleur <- meta.sub |> collapse::fsubset(Order == "Pleuronectiformes") |> nrow()
num.scorp <- meta.sub |> collapse::fsubset(Order == "Scorpaeniformes ") |> nrow()
num.clup <- meta.sub |> collapse::fsubset(Order == "Clupeiformes") |> nrow()
num.scomb <- meta.sub |> collapse::fsubset(Order == "Scombriformes") |> nrow()
num.perci <- meta.sub |> collapse::fsubset(Order == "Perciformes") |> nrow()

n.st.spec <- meta.sub |> collapse::fgroup_by(Species ) |> collapse::fsummarise(n.stocks = length(Stock)) 
n.st.spec <- n.st.spec[rev(order(n.st.spec$n.stocks)),]
num.cod <- meta.sub |> collapse::fsubset(Species == "morhua") |> nrow()
num.her <- meta.sub |> collapse::fsubset(Species == "harengus") |> nrow()
num.had <- meta.sub |> collapse::fsubset(Species == "aeglefinus") |> nrow()
num.sol <- meta.sub |> collapse::fsubset(Species == "solea") |> nrow()
num.pol <- meta.sub |> collapse::fsubset(Species == "virens") |> nrow()
num.plat <- meta.sub |> collapse::fsubset(Species == "platessa") |> nrow()

n.six <- meta.sub |> collapse::fsubset(Species %in% six.species) |> nrow()

med.lambda.F0 <- round(median(res.lambda.fin$lam.no.fish),digits=3)
med.r.F0 <- round(median(log(res.lambda.fin$lam.no.fish)),digits=3)*100
med.lambda.F <- round(median(res.lambda.fin$lam.fish),digits=3)
med.r.F <- round(median(log(res.lambda.fin$lam.fish)),digits=3)*100
mn.lambda.F <- round(mean(res.lambda.fin$lam.fish),digits=2)
mn.r.F <- round(mean(res.lambda.fin$lam.fish),digits=2)
# Mean - median...
res.lambda.fin$med.mn.diff.no.fish <- res.lambda.fin$lam.no.fish.med - res.lambda.fin$lam.no.fish.mn
res.lambda.fin$med.mn.diff.fish <- res.lambda.fin$lam.fish.med - res.lambda.fin$lam.fish.mn
# How many stocks had the mean higher than the median
n.mn.higher.med.no.fish <- length(unique(res.lambda.fin$Stock.short[res.lambda.fin$med.mn.diff.no.fish < 0]))
n.mn.higher.med.fish <- length(unique(res.lambda.fin$Stock.short[res.lambda.fin$med.mn.diff.fish < 0]))
low.mn.lambda.stocks <- unique(res.lambda.fin$Stock.short[res.lambda.fin$med.mn.diff.fish > 0])
n.nw.ice <- length(c(grep("NAFO",low.mn.lambda.stocks),grep("Faroe",low.mn.lambda.stocks)))
n.cod.low.mn <- length(grep("morhua",low.mn.lambda.stocks))
n.cod <- length(grep("morhua",Stocks))

@@

<!--chapter:end:tmp-01-analysis_no_age_correction.Rmd-->

# INTRODUCTION  {#intro}

Intro – life history paragraph, environmental variation paragraph, management ties paragraph, goals.

theoretical expectations related to productivity and life history for various species (e.g. herring vs. Cod) (ties to regional variation in lifetime reproductive rates and doubling time).

Evidence for specific species to make use of environmental conditions – boom in recruitment.

How our understanding of productivity is used in a management context – projections, reference points, risk, (ties directly to density dependence figure).  
Species-environment relationship – is it different from what we expect and how does that influence our capacity for management of exploited fishes.

The under appreciated influence of regional variation in determining recovery potential of exploited teleosts.  - can’t look at a single species and expect it will behave the same way.

Objectives
  - Quantify the life history variability of exploited marine teleosts
  - Explore regional and taxonomic correlates of life history variability
  - Based on the observed life history, quantify the recovery potential of marine teleosts in the absence of fishing

<!--chapter:end:tmp-02-intro.Rmd-->

# Methods {#meth}

## Data

This analysis used age structured fisheries data obtained from stock assessments for stocks in the Atlantic and Pacific Oceans [@charbonneauAgestructuredmarinefishdatabase2022; @charbonneauEffectsFishingMortality2022]. This repository had sufficient data to run the analysis for 77 stocks from NE Atlantic, NW Atlantic, and the North Pacific, this included stocks managed by International Council for the Exploration of the Sea (ICES; `r n.stock.ices`), NOAA (NOAA;`r n.stock.noaa`), and Fisheries and Oceans Canada (DFO; `r n.stock.dfo` stocks). The time series used in this analysis started as early as `r min.year` and the most recent data were from `r max.year`, the average length of the time series for the stocks included in this analysis was `r mn.ts.len` years and the number of years used ranged between `r min.ts.len` and `r max.ts.len` years. The stocks used represented `r num.orders` orders, with the most represented orders being the Gadiformes (`r num.gads` stocks), Pleuronectiformes (`r num.pleur` stocks) and Clupeiformes (`r num.clup` stocks). The other three orders had relatively few stocks included in the analysis, with `r num.perci` Perciform stocks, `r num.scomb` Scombriform stocks, and just `r num.scorp` Scorpaeniform stock.  The most common species in this analysis were Atlantic cod (*G. morhua*) and the majority of the stocks (`r n.six`) in this analysis were represented by one of six species (*G. morhua* (`r num.cod`), *C. harengus* (`r num.her`), *M. aeglefinus* (`r num.had`),  *P. virens* (`r num.pol`), *S. solea* (`r num.sol`), and *P. platessa* (`r num.plat`)).

These data included age-specific estimates of abundance, weight, biomass, maturity, natural mortality, and fishery removals (Table \@ref(tab:tab-meta)), these data were all required to calculate survival-at-age and fecundity-at-age and to estimate the annual productivity of these stocks. For the age structured analysis, the last age class was considered the 'terminal' age in the analysis, thus we implicitly assumed there was no contribution to productivity from older aged individuals. In addition, for the majority of the stocks the first age class was age 1 (`r n.age.1` stocks) while another `r n.age.2.plus` of the stocks have a first age of 2 or greater. The young age classes not included for these stocks (e.g. Age 0) consisted of immature individuals that do not contribute to the productivity of the stock; however it has been noted that the exclusion of immature age classes in these types of analyses may lead to a slight overestimate of productivity [@pardoMaximumIntrinsicRate2016].

## Life Table Reconstruction

The discrete time Euler-Lotka  equation was first proposed by Euler in 1760 (subsequently published by @moserGesetzeLebensdauer1839) and is a special case of the continuous time demographic model first proposed by Lotka in 1907 [@lotkaRelationsBirthRates1907; @sharpeProblemAgedistribution1911; @lotkaRelationBirthRate1918]:

$$
1 = \sum_{x=\alpha}^{\omega} \lambda^{-x} s_x f_x dx
$$

Where $x$ is the age, $\alpha$ is the age at first reproduction, $\omega$ is the maximum age, $\lambda$ is the discrete rate of population growth ($\lambda = e^r$), $s_x$ is the survivorship to age $x$, and $f_x$ is the fecundity of individuals of age $x$. Note that the discrete time Euler-Lotka equation can be shown to be the characteristic polynomial of the Leslie Matrix, where $f_0$ is the fecundity of the first age class with mature individuals [@leslieUseMatricesCertain1945, @caswellMatrixPopulationModels2001]. The available stock assessment data are used to calculate survival-at-age and fecundity-at-age to estimate $\lambda$, assuming that each stock represents a discrete population with individual dynamics.

<!-- \begin{align} -->
<!--   \begin{bmatrix} -->
<!--     f_0  & f_1 & f_2  & f_3    & ...          & f_{\omega-1} \\ -->
<!--     s_0  & 0   & 0    & 0      & ...          & 0 \\ -->
<!--     0    & s_1 & 0    & 0      & ...          & 0 \\ -->
<!--     0    & 0   & s_2  & 0      & ...          & 0 \\ -->
<!--     0    & 0   & 0    & \ddots & ...          & 0 \\ -->
<!--     0    & 0   & 0    & ...    & s_{\omega-2} & 0 \\ -->
<!--   \end{bmatrix} -->
<!-- \end{align} -->

### Fecundity-at-age

The number of spawners in year $y$ and age class $a$ ($SSN_{y,a}$) is the proportion of mature individuals ($Mat_{y,a}$) multiplied by the number of individuals in each age

$$
SSN_{y,a} = Mat_{y,a} N_{y,a}
$$
The Euler-Lotka equation requires fecundity to be the number of offspring produced by each female in each age class. Given the uncertainty of both fertilization and egg survival, using the number of eggs at a given length or size is inappropriate.  However, these data do have the number of recruits produced each year, and this can be combined with the spawning stock biomass to obtain an estimate of the number of recruits produced per kilogram of spawning biomass $RPS_{y}$, this is calculated as

$$
RPS_{y} = R_{y} SSB_{tot,(y-ra)} 
$$

Where $SSB_{yot,y-ra}$ is the total spawning stock biomass in year $y$. $R_{y}$ is the number of individuals in the youngest age class available for a specific stock (hereafter referred to as recruits) offset temporally by the age of the recruits. For example, if the age of the recruits ($ra$) is 2 years , we assume that the recruits in 2010 were produced by the spawning stock biomass in 2008 ($SSB_{tot,y-ra}$).  The number of recruits produced by each age class ($RPA_{y,a}$) can then be calculated by multiplying the spawner biomass in each age class $SSB_{y-ra,a}$ by the number of recruits produced per kilogram of SSB

$$
RPA_{y,a} =  SSB_{(y-ra),a} RPS_{y}
$$
This method does not account for any changes in reproductive success by age, each kilogram of spawning biomass contributes the same to the total total reproductive output, irrespective of their age. Finally, the age-specific fecundity ($Fec_{y,a}$) is calculated as number of offspring produced by each individual in the population. This is offset by one year, as the recruits observed in year $y+1$ enter into the Euler-Lotka formulation in year $y$.

$$
Fec_{y,a} =  \frac{RPA_{y+1,a}}{SSN_{(y-ra+1),a}}
$$
### Suvival-at-age

The number of animals surviving to each age depends on age-specific mortality rates. Commercially-exploited fishes experience fishing mortality (F) once they grow large enough to become vulnerable to the gear (reference). Fishing mortality combines with natural mortality (M) to give total mortality (Z) affecting each age class in each year. Varying assumptions can be made about natural mortality during stock assessments, where it may be an assumed value for each age (EXAMPLE), allowed to vary over time and by age (EXAMPLE), or input as a constant across ages (EXAMPLE).  Due to challenges with separability in typical stock assessment models, natural mortality is often assumed for each age class, in some assessments it varies by age and over time, in other assessments it does not.  Irrespective of this, the stock assessment used in this analysis all provide an estimate of natural mortality by age and year, this value is typically an instantaneous rate and is convert to a proportional natural mortality for the Euler-Lotka model.

The data also contains the number of fish harvested by the fishery each year and for each age. This can be combined with the assessed numbers at age to get an estimate of fishing mortality in each year for each age.

$$
FM_{y,a} =  \frac{Catch_{y,a}}{N_{y,a}}
$$
This is combined with the natural mortality to get a total mortality estimate ($Z_{y,a}$) for each age in each year

$$
Z_{y,a} =  FM_{y,a} + NM_{y,a}
$$
### Population Growth Rates $\lambda$

The fecundity, natural mortality, and fishing mortality estimates were then used to obtain an estimate of $\lambda$ for a given year from the Euler-Lotka equation and used to predict the abundance in the following year.

$$
N_{y+1} = \lambda_{y} N_{y}
$$
The Euler-Lotka model was fit using an iterative minimization process, where predicted abundance in the following year was compared to the abundance estimate from the stock assessment for the corresponding year ($N_{y+1}$). Where necessary, the annual age-specific fecundity and natural mortality estimates were adjusted until the $\lambda$ estimate from the Euler-Lotka equation resulted in an abundance time-series estimate ($N_{y+1}$) that was within 5% of the abundance estimates from the stock assessment (median difference was `r med.per.diff`). This tolerance ensured that our analyses reflected the same understanding of stock productivity, natural mortality and fecundity as in the current assessments; effectively this methodology treats the abundance estimates from the stock assessments as the 'truth'.

The life table reconstruction resulted in stock-specific $\lambda$, fecundity, and natural mortality estimates for each year in which data were available accounting for the age structure and life history characteristics of each stock. In other words, we estimated the realized population growth rate using the changes in the population size of the stocks from year to year (realized population growth rates), based on the amount of mortality experienced and annual reproductive success. The realized $\lambda$
included fishing mortality  ($\lambda_{F>0}$), although this method enabled the theoretical population growth rate in the absence of fishing  ($\lambda_{F=0}$) could also be calculated directly fomr the annual the natural mortality and fecundity estimates.

<!-- For consistency in the Euler-lotka estimates across all the stocks, the first age used in the analysis was 'age 0' (which is simply an estimate of abundance of fish in the year they were born). For some stocks this information is not available (see Table X, **possibly in an Appendix where we have meta-data**). For the stocks in which age 0 information is unavailable, an estimate of the abundance of the younger age classes were back calculated using average natural mortality provided for the youngest age class available. -->

## Density Dependence

Life table analyses make no assumptions about population regulation and thus do not incorporate a density-dependent relationship between spawners and recruits (Reference?). However, the data itself could be explored for evidence of density-dependence by modelling the estimated annual population growth rate against the proportion of maximum abundance observed that year for each stock. This was done twice, once using  ($\lambda_{F=0}$) and once using ($\lambda_{F>0}$).  Negative density dependence is observed when $\lambda$ increases as abundance declines, positive density dependence occurs when $\lambda$ declines with increasing abundance, while density independence is observed when $\lambda$ is not impacted by changes in abundance.

Generalized Additive Mixed Models were used to allow for some non-linearity in the relationships (CITE MCGV and WOOD), the overall effect and the effect for each stock was modelled as

\begin{align}
\lambda_{i,j} \sim N(0,\sigma^2_{i,j}) \\
E(\lambda_{i,j}) = \mu_{i,j} \\
log(\mu_{i,j}) =  Prop_{i} + Stock_{j}  \times Prop_{i} \\
Stocks_{i,j} \sim N(0,\sigma^2_{Stocks})
\end{align}

The interaction was modeled using the *fs* basis using the mgcv package in R, this treats the smooth curves as random effects and is used when a factor has a large number of levels (CITE!).  Various alternative modelling methods were explored (various model formulations, alternative error distributions, etc.) but all models had similar challenges overcoming either the underlying non-normality or the heteroskedasticity of the data. As such, the uncertainty surrounding the model output should be treated with caution, while we discuss the general trends this model provides we attempted to avoid discussion of the statistical significance of the trends.

## Demographic Parameters

There are numerous demographic parameters that can be derived from the outputs of the life table reconstruction. Here we focus on two that are strongly linked to recovery potential: lifetime reproductive success and generation time. Lifetime reproductive success describes the average number of spawners produced by one female spawner throughout her lifetime

## Lifetime Reproductive Success
